Nutritional Deficits and the Quantity-Quality Trade-off: Evidence from an Exogenous Fertility Shock in Low-Income Urban Settings in the Philippines

### 2025-06-10 docs/paper.Rmd
start RRL section with theoretical background

---
title: "Nutritional Deficits and the Quantity-Quality Trade-off: Evidence from an Exogenous Fertility Shock in Low-Income Urban Settings in the Philippines"
short: "The Beckerian Quantity-Quality Trade-off in the Philippine Context"
journal: "" # AER, AEJ, PP, JEL
month: "`r format(Sys.Date(), '%m')`"
year: "`r format(Sys.Date(), '%Y')`"
vol: 1
issue: 1
jel:
  - J13  # Fertility; Childbearing; Family Planning; Population Policy
  - I15  # Health and Economic Development
  - O15  # Human Resources; Human Development; Income Distribution; Migration
keywords:
  - fertility shocks
  - child nutrition
  - quantity-quality trade-off
  - urban poverty
  - Philippines
author:
  - name: Erika Salvador
    firstname: Erika
    surname: Salvador
    email: esalvador28@amherst.edu
    affiliation: Amherst College
  - name: Caroline Theoharides
    firstname: Caroline
    surname: Theoharides
    email: ctheoharides@amherst.edu
    affiliation: Amherst College
acknowledgements: |
  This research was made possible by the Schupf Fellowship at Amherst College. I am indebted to my faculty adviser, Professor Caroline Theoharides, for her mentorship and support throughout the development of this project. I also wish to thank Faculty Director Professor Amelie Hastie and the campus partners whose institutional support enabled this work. I am further grateful to the Departments of Economics and Mathematics & Statistics for their academic support. The views expressed and any errors contained herein are entirely my own responsibility.
abstract: |
  This paper examines whether increased fertility affects early-life nutritional outcomes in low-income urban households. I exploit a natural experiment created by a 1990 policy in Manila, Philippines, which banned modern contraceptives from city-run health facilities. Using a difference-in-differences framework and nationally representative data from the Philippine Demographic and Health Surveys, I estimate the reduced-form impact of the policy on child height-for-age and weight-for-height. [Will add more after data analysis]
output: rticles::aea_article
bibliography: references.bib
---

@@

The analysis proceeds in three stages. I first replicate existing work \citep{dumas2010fertility} to confirm that the contraceptive ban led to an increase in fertility among affected women. I then estimate reduced-form effects of policy exposure on nutritional outcomes for children under five years of age. Finally, I examine heterogeneity in effects across subsamples defined by maternal education, household wealth, and access to prenatal care. These dimensions serve as proxies for household resource availability and capacity to buffer the nutritional consequences of fertility increases.

@@

>>>>

---
title: "Nutritional Deficits and the Quantity-Quality Trade-off: Evidence from an Exogenous Fertility Shock in Low-Income Urban Settings in the Philippines"
short: "The Q-Q Trade-off in the Philippine Context"
journal: "" # AER, AEJ, PP, JEL
month: "`r format(Sys.Date(), '%m')`"
year: "`r format(Sys.Date(), '%Y')`"
vol: 1
issue: 1
jel:
  - J13  # Fertility; Childbearing; Family Planning; Population Policy
  - I15  # Health and Economic Development
  - O15  # Human Resources; Human Development; Income Distribution; Migration
keywords:
  - fertility shocks
  - child nutrition
  - quantity-quality trade-off
  - urban poverty
  - Philippines
author:
  - name: Erika Salvador
    firstname: Erika
    surname: Salvador
    email: esalvador28@amherst.edu
    affiliation: Amherst College
  - name: Caroline Theoharides
    firstname: Caroline
    surname: Theoharides
    email: ctheoharides@amherst.edu
    affiliation: Amherst College
acknowledgements: |
  This research was made possible by the Schupf Fellowship at Amherst College. I am indebted to my faculty adviser, Professor Caroline Theoharides, for her mentorship and support throughout the development of this project. I also wish to thank Faculty Director Professor Amelie Hastie and the campus partners whose institutional support enabled this work. I am further grateful to the Departments of Economics and Mathematics & Statistics for their academic support. The views expressed and any errors contained herein are entirely my own responsibility.
abstract: |
  This paper examines whether increased fertility affects early-life nutritional outcomes in low-income urban households. I exploit a natural experiment created by a 1990 policy in Manila, Philippines, which banned modern contraceptives from city-run health facilities. Using a difference-in-differences framework and nationally representative data from the Philippine Demographic and Health Surveys, I estimate the reduced-form impact of the policy on child height-for-age and weight-for-height. [Will add more after data analysis]
output: rticles::aea_article
bibliography: references.bib
---

@@

The analysis proceeds in three stages. I first replicate existing work \citep{dumas2019sex} to confirm that the contraceptive ban led to an increase in fertility among affected women. I then estimate reduced-form effects of policy exposure on nutritional outcomes for children under five years of age. Finally, I examine heterogeneity in effects across subsamples defined by maternal education, household wealth, and access to prenatal care. These dimensions serve as proxies for household resource availability and capacity to buffer the nutritional consequences of fertility increases.

@@

\section{Review of Related Literature}

\subsection{Theoretical Background}

The quantity–quality (Q–Q) trade-off is a core idea in the economics of fertility. It formalizes the constraint that when parents face a fixed income, they must allocate limited resources between the number of children they have and the amount invested in each child's development. In this framework, “quality” refers to any investment that raises a child’s human capital—such as nutrition, education, healthcare—while “quantity” refers simply to the number of children. Because these investments are resource-intensive, increasing one often comes at the expense of the other.

Becker’s foundational model \citep{becker1960economic} formalizes this as a constrained optimization problem. A household maximizes utility over the number of children \(n\), child quality \(q\), and consumption \(c\):

\[\max_{n,q,c} \quad U(n, q, c)\]

subject to the budget constraint:

\[I = c + n(p_n + p_q q),\]

where \(I\) is household income, \(p_n\) is the fixed cost per child (e.g., food, shelter), and \(p_q\) is the marginal cost of investing in child quality. The budget constraint makes the trade-off explicit: increasing \(n\) reduces how much can be spent on \(q\), and vice versa.

### 2025-06-10 docs/paper.tex
override AER default to display SCHUPF as issue name
start RRL section with theoretical background

% AER-Article.tex for AEA last revised 22 June 2011
\documentclass[2025 Schupf Fellowship Project]{AEA}

@@

\title{Nutritional Deficits and the Quantity-Quality Trade-off: Evidence
from an Exogenous Fertility Shock in Low-Income Urban Settings in the
Philippines}
\shortTitle{The Beckerian Quantity-Quality Trade-off in the Philippine
Context}
% \author{Author1 and Author2\thanks{Surname1: affiliation1, address1, email1.
% Surname2: affiliation2, address2, email2. Acknowledgements}}

@@

>>>>

% AER-Article.tex for AEA last revised 22 June 2011
\documentclass[]{AEA}

@@

\title{Nutritional Deficits and the Quantity-Quality Trade-off: Evidence
from an Exogenous Fertility Shock in Low-Income Urban Settings in the
Philippines}
\shortTitle{The Q-Q Trade-off in the Philippine Context}
% \author{Author1 and Author2\thanks{Surname1: affiliation1, address1, email1.
% Surname2: affiliation2, address2, email2. Acknowledgements}}

@@

\section{Review of Related Literature}

\subsection{Theoretical Background}

The quantity--quality (Q--Q) trade-off is a core idea in the economics
of fertility. It formalizes the constraint that when parents face a
fixed income, they must allocate limited resources between the number of
children they have and the amount invested in each child's development.
In this framework, ``quality'' refers to any investment that raises a
child's human capital---such as nutrition, education, healthcare---while
``quantity'' refers simply to the number of children. Because these
investments are resource-intensive, increasing one often comes at the
expense of the other.

Becker's foundational model \citep{becker1960economic} formalizes this
as a constrained optimization problem. A household maximizes utility
over the number of children \(n\), child quality \(q\), and consumption
\(c\):

\[\max_{n,q,c} \quad U(n, q, c)\]

subject to the budget constraint:

\[I = c + n(p_n + p_q q),\]

where \(I\) is household income, \(p_n\) is the fixed cost per child
(e.g., food, shelter), and \(p_q\) is the marginal cost of investing in
child quality. The budget constraint makes the trade-off explicit:
increasing \(n\) reduces how much can be spent on \(q\), and vice versa.
